Evaluation of a Cache-Oblivious Data Structure [draft-1]

### 2008-05-31 paper/paper.tex
Updated the paper (updated results included).

@@

Finally, the Bender set uses a complete binary tree as an index data structure
to efficiently find the successor element for a given value. The tree is
stored in memory in van Emde Boas layout (as described by Prokop) to allow
efficient cache-oblivious traversal.

In the tree, the leaf nodes (when numbered from left to right) store the same values
as the sparse array (including empty values).
Interior nodes store the maximum value of their two children (or an empty value,
if both children store an empty value).
This tree can be used in a similar way as a binary search tree: by examining the
two children of a node, we can determine whether a searched-for value lies in the
left or right subtree.

@@

Therefore, we decided to measure wall clock time. To deal with variations
across multiple executions, each experiment is run five times, and the median
value is used for further analysis.

@@

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table.
The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the (fixed) size of the index.
Although for this case the hash table with an index size of 1,000,000 is
slightly faster, we will use an index size of 10,000,000 for further
experiments, as these cases require more values to be stored.

@@

>>>>

The lack of unrealistic requirements combined with the promise of cache-efficiency accross multiple levels of the memory hierarchy, makes it an interesting model for the development of low-maintenance, high-performance data structures for use in real-world applications.

@@

Finally, the Bender set uses a complete binary tree as an index data structure
to efficiently find the successor element for a given value. The tree is
stored in memory in van Emde Boas layout to allow cache-oblivious
traversal. This layout is named after the data structure described in
\cite{vanemdeboas1976dai} and determines an order in which the nodes of a
tree can be stored in a contiguous block of memory, in such a way that traversing
a tree with $N$ nodes from root to leaf requires only $O(\log_B N + 1)$ pages to be
fetched from memory.

In van Emde Boas lay-out, which is defined recursively, a tree with 
$N$ nodes is divided in half vertically, creating a small subtree at the
top and a number of subtrees (approximately $\sqrt{N}$) below. Each subtree
has a size of approximately $\sqrt{N}$ nodes and will be stored in a
contiguous block of memory.

In Figure \ref{fig-van-emde-boas} a binary tree is shown, with the nodes
labeled with their positions according to the van Emde Boas layout.
In this example, only two levels of recursion are needed. If each page
stores three nodes, the path from node 1 to node 12 (or any other leaf node)
requires visits only two pages.

\begin{figure}
\centering
\includegraphics[width=83mm]{vanemdeboas}
\caption{A binary tree in van Emde Boas layout}
\label{fig-van-emde-boas}
\end{figure}

In the index tree used for the Bender set, the leaf nodes store the same values
as the sparse array (including empty values).
Interior nodes store the maximum value of their two children (or an empty value,
if both children store an empty value).
This tree can be used in a similar way as a binary search tree: by examining the
two children of a node, we can determine whether a searched-for value lies in the
left or right subtree.

@@

Therefore, we decided to measure wall clock time. To deal with variations
across multiple executions, each experiment is run seven times, and the median
value is used for further analysis.

@@

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table.
The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the (fixed) size of the index.
We will use the hash table with an index size of 10,000,000 for further
experiments, as it performed best in this case, and seems suitable for other
cases (which require storing more values) as well.

@@

The increased memory requirements of the Bender set can in part be explained by
our implementation of variable-length values, which wastes some space by storing
them in fixed-size slots. This overhead should be less than 25\% on average
(and at most 50\% in the worst case).
